---
title: "LLMs for Journalism with R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Your Assignment

Throughout these exercises, you are investigating attorney discipline in Maryland. The Maryland Office of Bar Counsel maintains public records of [sanctions](https://www.courts.state.md.us/attygrievance/sanctions) against attorneys — disbarments, suspensions and reprimands. Two years of that data (`sanctionsfy25.txt` and `sanctionsfy26.txt`) are already in this repository.

By the end of these exercises, you'll have turned those unstructured text files into a structured, analyzable dataset and can start asking reporting questions — using LLMs to do in minutes what would otherwise take hours by hand.

## Text Models

The `ellmer` package lets us have conversations with LLMs, ask them to summarize text and extract structured data. Here's how to fetch a published story and summarize it in a few lines of code:

```{r summarize-article}
library(ellmer)
library(rvest)

url <- "https://cnsmaryland.org/2025/08/06/many-towns-unprepared-for-derailments-hazmat-spills-that-can-trigger-fires-evacuations/"
page <- read_html(url)
story_text <- page |> html_elements("article p") |> html_text2() |> paste(collapse = "\n")

chat <- chat_groq(model = "meta-llama/llama-4-scout-17b-16e-instruct")
chat$chat(paste("Summarize this story in 3 paragraphs:\n\n", story_text))
```

Run this and read the output. Now imagine doing this for 300 stories at once, or 300 PDFs.

## Restructuring Information

Text models can do more than summarize; they can restructure information to make it more useful. One of `ellmer`'s best features is **structured data extraction** -- you can define the exact shape of the data you want and the LLM will return it as an R data frame.

We'll take a document describing sanctions of Maryland attorneys (the `sanctionsfy25.txt` file in this repository) and ask the LLM to turn unstructured text into structured data:

```{r structured-extraction}
library(jsonlite)

sanctions_text <- readLines("sanctionsfy25.txt") |> paste(collapse = "\n")

chat <- chat_groq(model = "meta-llama/llama-4-scout-17b-16e-instruct", params = params(max_tokens = 8192))

response <- chat$chat(paste(
  "Extract attorney sanctions from this text as a JSON array.",
  "Each object should have these keys: name, sanction, date (in yyyy-mm-dd format), description.",
  "Return ONLY the JSON array, no other text.\n\n",
  sanctions_text
))

result <- fromJSON(response)

print(result)
```

Notice that `result` is a data frame you can work with in R -- filter, sort, export to CSV, whatever you need.

But how good is it? Open `sanctionsfy25.txt` and spot-check the results. Count the sanctions in the original text and compare that to `nrow(result)`. Pick two or three rows and verify the name, date, and sanction type against the source. Did the LLM miss any entries? Did it hallucinate any that aren't in the document? Did it get any dates wrong?

**Discussion: How many sanctions are in the original text vs. the data frame? Which fields were most accurate — names, dates, sanction types, or descriptions? What kind of errors did you find, and how would you catch them at scale?**

Now it's your turn: process `sanctionsfy26.txt` the same way (or change something!). Use the code above as a starting point, and do the same validation — check the count and spot-check a few rows against the source text.

```{r your-structured}

sanctions_text_fy26 <- readLines("sanctionsfy26.txt") |> paste(collapse = "\n")

chat <- chat_groq(model = "meta-llama/llama-4-scout-17b-16e-instruct")

response_fy26 <- chat$chat(paste(
  "Extract attorney sanctions from this text as a JSON array.",
  "Each object should have these keys: name, sanction, date (in yyyy-mm-dd format), description.",
  "Return ONLY the JSON array, no other text.\n\n",
  sanctions_text_fy26
))

result_fy26 <- fromJSON(response_fy26)

print(result_fy26)
```

**Discussion: Look at both data frames. What's the most surprising difference between the two years? If an editor asked you to pitch a story from this data, what would you say?**

## Named Entity Recognition

Extracting entities (people, organizations, places) from text is a common journalism task. With `ellmer`, you can define exactly what you want to extract:

```{r ner}

article_text <- "FRANKLIN, Jamel R. – Disbarment by Consent on December 23, 2024,
effective immediately, for committing a criminal act that reflects adversely on his
honesty, trustworthiness, or fitness as an attorney; and engaging in conduct involving
dishonesty, fraud, deceit, or misrepresentation. The Respondent pled guilty to one count
of perjury in violation of Maryland Code, Criminal Law Article § 9-101; and one count of
theft over $100,000.00 in violation of Maryland Code, Criminal Law Article § 7-104."

chat <- chat_groq(model = "meta-llama/llama-4-scout-17b-16e-instruct")

response <- chat$chat(paste(
  "Extract all named entities from this text as a JSON array.",
  "Each object should have these keys: name, type (one of: person, organization, place, money, date, law), context (brief context for how the entity appears).",
  "Return ONLY the JSON array, no other text.\n\n",
  article_text
))

entities <- fromJSON(response)

print(entities)
```

**Discussion: The LLM extracted entities using categories we defined. What did it miss or get wrong? What categories would you add if you were building a database to track attorney misconduct over time?**

## Vision Models

We can use models that understand images to extract information from documents, charts and photos. Let's test that using an image from a Maryland physician discipline report. The image `md_doc.png` is already in this repository - take a look at it first.

```{r vision}

chat <- chat_google_gemini(model = "gemini-2.5-flash")
chat$chat(
  content_image_file("md_doc.png"),
  "What is the license number from this image?"
)
```

You should get something like: "The license number is D25922."

Now it's your turn: find an image *that you would be ok showing to your grandmother* and make sure it's a PNG, GIF or JPG file. Drag it into your llm-playground repository so you can see it in your list of files, then change the code below to refer to your image and ask a different question. How did the LLM do?

```{r your-vision}
library(ellmer)

chat <- chat_google_gemini(model = "gemini-2.5-flash")
chat$chat(
  content_image_file("YOUR_FILE"),
  "YOUR QUESTION"
)
```

**Discussion: What did the LLM get right about your image? What did it get wrong or miss? Would you trust this output enough to use it in a story without checking?**

## Conversations

One powerful feature of `ellmer` is that chat objects remember the conversation. You can ask follow-up questions without repeating context. Here we'll use the sanctions data we extracted earlier to explore it the way a reporter would:

```{r conversations}

# Using the result data frame from the structured extraction above
sanctions_context <- paste(result$name, result$sanction, result$date, sep = " | ", collapse = "\n")

chat <- chat_groq(model = "meta-llama/llama-4-scout-17b-16e-instruct")
chat$chat(paste("Here are Maryland attorney sanctions from FY2025:\n\n", sanctions_context))
chat$chat("Which type of sanction appears most often, and what does that suggest about attorney conduct?")
chat$chat("If you were reporting on attorney discipline for a news organization, what would be the most newsworthy finding here?")
chat$chat("What additional records or sources would you seek out to report this story?")
```

Each response builds on the previous ones — the LLM retains context across all three questions without you having to repeat the sanctions list each time.

You don't want to use this technique to _replace_ reporting and your own judgment. You should think of AI as your assistant: it tries to be helpful but sometimes doesn't know what it's talking about and doesn't always know what you know.

## Batch Processing with tidyllm

Everything we've done so far processes one thing at a time. But what if you have 50 articles to classify, or 200 documents to extract names from? Sending them one by one can be slow and expensive. The `tidyllm` package solves this with **batch processing** -- you submit many prompts at once and the LLM provider processes them in the background, often at a 50% discount.

`tidyllm` uses a different style than `ellmer`. Instead of a chat object, you build a pipeline: create a message, pipe it to a provider, extract the reply. Here's a single prompt to see how it works:

```{r tidyllm-basics}
library(tidyllm)

llm_message("What is the capital of Maryland?") |>
  chat(groq(.model = "meta-llama/llama-4-scout-17b-16e-instruct")) |>
  get_reply()
```

The real power is batch processing. Now that you have the FY2025 sanctions as a structured data frame, you can classify all the misconduct descriptions at once instead of one at a time:

```{r batch-send}
library(tidyverse)
library(glue)

# Using the result data frame from the structured extraction above
prompts <- glue(
  "Classify this attorney misconduct description into exactly one category: Financial, Communication, Criminal, Conflict of Interest, Other.
   Respond with only the category name, no punctuation.
   Description: {description}",
  description = result$description
)

tasks <- map(prompts, llm_message)

batch_result <- tasks |>
  send_batch(groq(.model = "meta-llama/llama-4-scout-17b-16e-instruct", .temperature = 0))

saveRDS(batch_result, "my_batch.rds")
```

The batch is now processing in the background. You can check on it and retrieve results:

```{r batch-fetch}
readRDS("my_batch.rds") |>
  check_batch(groq())

conversations <- readRDS("my_batch.rds") |>
  fetch_batch(groq())

categories <- map_chr(conversations, get_reply)

tibble(name = result$name, sanction = result$sanction, misconduct_category = categories)
```

This is a game-changer for journalism: you could classify an entire year of city council agendas, extract names from hundreds of campaign finance reports, or categorize thousands of police incident reports -- all in a single batch. Providers like Claude and OpenAI offer batch processing at half the normal price.

**Discussion: Did the LLM miscategorize any of the sanctions? Pick one you'd classify differently and explain why. What does that tell you about relying on automated classification for a published story?**

## Putting It Together

You've extracted structured data from sanctions records, run entity recognition on the text, used multi-turn conversation to explore the data, and classified misconduct types in bulk. Now combine both years into one dataset and compare:

```{r capstone}

sanctions_text_fy25 <- readLines("sanctionsfy25.txt") |> paste(collapse = "\n")
sanctions_text_fy26 <- readLines("sanctionsfy26.txt") |> paste(collapse = "\n")

chat25 <- chat_groq(model = "meta-llama/llama-4-scout-17b-16e-instruct", params = params(max_tokens = 8192))
response25 <- chat25$chat(paste(
  "Extract attorney sanctions from this text as a JSON array.",
  "Each object should have these keys: name, sanction, date (in yyyy-mm-dd format), description.",
  "Return ONLY the JSON array, no other text.\n\n",
  sanctions_text_fy25
))
fy25 <- fromJSON(response25) |> mutate(fiscal_year = "FY2025")

chat26 <- chat_groq(model = "meta-llama/llama-4-scout-17b-16e-instruct", params = params(max_tokens = 8192))
response26 <- chat26$chat(paste(
  "Extract attorney sanctions from this text as a JSON array.",
  "Each object should have these keys: name, sanction, date (in yyyy-mm-dd format), description.",
  "Return ONLY the JSON array, no other text.\n\n",
  sanctions_text_fy26
))
fy26 <- fromJSON(response26) |> mutate(fiscal_year = "FY2026")

all_sanctions <- bind_rows(fy25, fy26)

all_sanctions |>
  count(fiscal_year, sanction, sort = TRUE) |>
  pivot_wider(names_from = fiscal_year, values_from = n, values_fill = 0)
```

**Discussion: Before you write anything, validate: does the total number of sanctions per year in `all_sanctions` match what you counted in the original text files? If not, what went wrong — and would a reader ever know? Now write one lede sentence that captures the most newsworthy change, and say who you'd call first to report it out.**

## How You Could Use AI

**Discussion: Pick a beat or topic you cover (or want to cover). Describe one specific task — not a goal, a task — where these tools would change how you report. What would you feed in, and what would you get out?**

## Where This Is Heading

The techniques you've practiced here — summarization, structured extraction, entity recognition, conversation, and batch classification — are building blocks. Individually they save time. Combined at scale, they open up entirely new kinds of reporting.

One example: [Beat Book Builder](https://github.com/NewsAppsUMD/beat-book) is a tool built by journalism students at Maryland that takes a corpus of news articles, automatically discovers topic clusters across them using embeddings and clustering, and then uses an AI agent to interview the reporter and generate a tailored beat book — a practical reporting guide with story ideas, source suggestions, and context. It's the logical extension of everything in this notebook: structured extraction feeds the ingest pipeline, NER surfaces the entities worth tracking, batch processing handles the scale, and a multi-turn conversation with the agent produces the final output.

The difference between these exercises and that tool is mostly engineering. The journalism thinking — deciding what to extract, what questions to ask, what makes a finding newsworthy — is still yours.

